`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-05-19 manuscript/manuscript.tex
render

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Habitat International}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi},
  pdfauthor={Elizabeth Tilley; Hope Chilunga; Jonathan Kwangulero; Lars Schöbitz; Saloni Vijay; Heiko Heilgendorff; Marc Kalina},
  pdfkeywords={waste management, healthcare waste, trash burning, air
quality, Malawi, urbanisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangerlo@gmail.comm} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[1]{Saloni Vijay%
%
}
 \ead{svijay@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{heiko.heilgendorff@gmail.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus\footnote{Sharps are handled and disposed of separately.}
\citep{kalina2020this}. However, the vast majority of the deposited
waste is not incinerated, as a new incinerator provided by the Ministry
of Health in late 2019, is only able to handle a small percentage of the
hospital's waste. Furthermore, the incinerator is frequently inoperable,
under repair, or without power\footnote{There was an older incinerator
  on the grounds which failed in 2017. Between 2017 and 2021, when the
  new incinerator was commissioned, nothing was incinerated at all.}.
Rather, the waste is openly burned, in a constantly smouldering pile to
the side of the incinerator building, sending up the aforementioned
clouds of smoke, which choke the grounds
(Figure~\ref{fig-open-burning}). Innumerable other, smaller, fires,
scattered across the QECH campus (Figure~\ref{fig-map-sensors})
contribute their part, as grounds and maintenance staff habitually burn
leaves and other garden refuse. Furthermore, caregivers\footnote{Patients
  are required to bring a caregiver for in-patient hospitals stays,
  usually a family member, who assist with feeding, bathing, and all
  other non-medical services.} who reside separately on hospital
grounds, cook for themselves throughout the day and night using biomass
(charcoal or wood), and burn their own domestic waste. As a result, air
quality at the hospital is a constant source of discussion, with
patients, staff, and caregivers struggling to cope, when just breathing
is, as the quote\footnote{ID5, 2019-11-18} used in the title of this
article, `unbearable.'

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trashing
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM\textsubscript{2.5}
anthropogenic emissions, it is of particular concern in Sub-Saharan
African (SSA) cities, which are home to 19 of the world's 50 biggest
dumpsites (United Nations Development Programme, 2021). Particulate
matter, or PM, is not specific to trash burning, but rather the product
of any type of incomplete combustion and is one of the most commonly
measured indicators for air quality. PM is measured in terms of its
particle size and generally classified into PM with particle sizes less
than 10 µm (PM\textsubscript{10}) and those with sizes less than 2.5 µm
(PM\textsubscript{2.5}), while one way of measuring the severity of air
pollution is by expressing the mass of PM per volume of air (µg
PMx/m\textsuperscript{3} air). PM has been linked to multiple negative
health outcomes. Amongst particulate air pollution, PM\textsubscript{10}
and PM\textsubscript{2.5} are of particular concern as they can
penetrate deep into the lungs, and their exposure has been associated
with asthma, heart disease, heart failure, stroke, and cancer
\citep{mohapatra2014effect, xing2016impact}. As a result, the health
impacts of ambient air particle pollution may be significant, especially
for the most vulnerable. For instance, the \citet{ihme2019ambient}
estimates PM related air pollution resulted in 4.14 million deaths
worldwide (in 2019), of which, 394,000 occurred in Africa. Yet, despite
its importance, the air quality and health picture in Africa is
worryingly incomplete. For instance, the WHO only collects data from 10
African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

\begin{itemize}
\item
  \textbf{Lions Sight First Eye Hospital}, often shortened to
  \textbf{Lions Sight}, is the largest eye hospital in Malawi and serves
  as the main teaching eye hospital for the Kamuzu University of Health
  Sciences (formerly College of Medicine). It is staffed by 5
  consultants and a team of clinical officers and nurses. It provides a
  mix of public (no cost) and private (at cost) services.
\item
  \textbf{The Blantyre Malaria Project (BMP)}, established by Professors
  Terrie Taylor and Malcolm Molyneux, has carried out clinical research
  and patient care in the area of paediatric malaria, specifically
  cerebral malaria, for more than 30 years\footnote{BMP is credited with
    the development of the Blantyre Coma Score, which is used widely.}.
  Its research infrastructure includes an administrative team, an
  inpatient research ward, an MRI centre, and an outpatient research
  clinic in Ndirande township and in other districts outside of
  Blantyre.
\item
  \textbf{Ward 6B} is a male ward for trauma and orthopaedic patients.
  It was established as part of the original hospital design. It is a
  60-bed ward with a dedicated nursing station and treatment rooms.
\item
  \textbf{The Blantyre Lighthouse Trust Clinic}, also known as
  \textbf{Umodzi Family Centre} or simply, the \textbf{Lighthouse
  Clinic}, is one of four operating across Malawi. The clinics work in
  collaboration with the Ministry of Health (MOH) to provide integrated
  HIV testing, treatment and care for people living with HIV.
\item
  The \textbf{Moyo Nutritional Rehabilitation and Research Unit}, known
  on campus as \textbf{uMoyo}, is a 57-bed nutritional rehabilitation
  unit (NRU) for treating infants and children with severe malnutrition
  and acute illnesses. It is one of 104 operational NRUs in Malawi. It
  also has an outpatient therapeutic feeding program (OTP) for children
  with malnutrition who can be treated outside of the hospital.
\item
  \textbf{Mercy James Institute for Pediatric Surgery and Intensive Care
  (MJC)}, shortened on campus to \textbf{Mercy James}, was opened in
  2017 by the NGO Raising Malawi, which was founded by Madonna. With 3
  operating rooms and 50 beds, it is the first and only first dedicated
  paediatric surgery hospital in Malawi. Prior to its opening, QECH had
  fewer than 10 intensive care beds.
\item
  The \textbf{Administration Building} is the operational hub of the
  campus, containing mostly offices and meeting rooms.
\item
  When a patient arrives at QECH, she or he will usually arrive with a
  guardian: someone to cook for them, buy medicine, do their laundry,
  and help them bathe. The \textbf{Guardian Shelter} is a
  gender-separated concrete floor shelter for sleeping along with a
  cooking pavilion and toilet/shower block are maintained by a the local
  NGO Chira fund. Though simple, the facility is secure and offers one
  of the only free/accessible toilet facilities for visitors on the QECH
  campus.
\end{itemize}

@@

Despite its national prominence, and international reputation for
academic medical research, Queen's is plagued by constant shortages of
human and material resources. Yet, these shortages are not shared
equally across the space, with a sharp division in cleanliness,
resources, and quality of care between public wards, supported by the
Malawi Ministry of Health, and the private wads with international
funding, such as the aforementioned Mercy James Institute. For staff,
these shortages have manifested in poor salaries, missed pay cheques,
failing infrastructure and limited supplies \citep{kalina2020this}.
Regarding waste management, these shortages have manifested in a chronic
shortage of black bags, waste containers and cleaning staff. As a
result, due to a lack of bags and storage containers, there is minimal
segregation of waste, and highly infectious items (bandages, gloves,
syringes, and other sharps) are disposed of along with non-hazardous
waste (food, newspapers, packaging). As a result, rather than
facilitating different disposal pathways for hazardous and non-hazardous
waste, all waste is mixed, necessitating more burning than would be
necessary if non-hazardous waste was diverted, and contributing to
workplace risk, with reports of accidental exposure to potentially
infected waste, including accidental pokes from improperly disposed of
sharps, common amongst hospital staff. Regardless, separation ends at
the point of collection. The hospital is not serviced by regular
municipal waste collection, and as a result, nearly all of the waste
produced by the hospital, hazardous and non-hazardous, has to be
disposed of on-site, either within the limited capacity of the hospital
incinerator, or more commonly, through open burning. As a result, the
constantly smouldering pile of waste puts forth a continuous cloud of
grey smoke, which mingling with the dozens of other fires on the
grounds, from cooking fires and burning garden waste, blankets Queen's
in a permanent cloud of foul smelling haze.

@@

The PM monitoring device centres around a Raspberry Pi 3 Model B
single-board computer complete with an operating system and storage
space. The added advantage of storage space is that the data are not
lost if the internet connection breaks down. Because the Pi can only
read digital signals, we needed to include an analogue-to-digital (ADC)
converter. The amount of dust is measured by a Nova PM SDS011 High
Precision Laser sensor. This sensor measures particles at 2.5 and 10 µm
in diameter in mg/m\textsuperscript{3}. A list of the specific hardware
and software components can be found in Appendix B.

@@

Raw data and analysis-ready processed data is available as an R package
at \citet{R-qechairquality}. The data underlying the tables and figures
of this manuscript are contained in a repository alongside a
reproducible document that contains the analysis code and the written
narrative of the manuscript {[}TODO: manuscript software citation{]}.

@@

\hypertarget{tbl-peaks}{}
\begin{longtable}[]{@{}lrrr@{}}
\caption{\label{tbl-peaks}Peaks for both PM\textsubscript{2.5} and
PM\textsubscript{10} are the number of data points above the WHO interim
target 1 (annual) which is the least stringent; the number of
observations recorded (count) is provided for reference.}\tabularnewline
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endfirsthead
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endhead
Administration & 10900 & 171 & 0 \\
Guardian Shelter & 11032 & 4529 & 0 \\
Lighthouse Clinic & 11472 & 4365 & 0 \\
Lions Sight & 11749 & 91 & 0 \\
Malaria & 12031 & 617 & 0 \\
Mercy James & 12193 & 736 & 0 \\
Ward 6B & 11861 & 1113 & 0 \\
uMoyo & 10637 & 703 & 0 \\
\bottomrule()
\end{longtable}

Over the course of the two-month study, the hazardous limits for both
parameters were exceeded at all locations except for the Lions Sight
hospital. The limits were exceeded fewer than 50 times at five of the
locations, and only the monitors at the Lighthouse Clinic and the
Guardian Shelter recorded more than 50 instances above the hazardous
limits. At the Lighthouse Clinic, the PM\textsubscript{10} limit was
exceeded almost twice as often as the PM\textsubscript{2.5} limit, while
at the Guardian Shelter, the reverse is true. Because the Lighthouse
Clinic is within 50 m of the incinerator, the number and predominance of
PM\textsubscript{10} peaks is characteristic of incomplete combustion
and dust that is typical for the area and the open burning that occurs
there. The residents at the Guardian Shelter however are mostly cooking
within contained clay stoves that have been designed to improve
combustion of the coal and wood that is burned within them. The fewer
number of PM\textsubscript{10} peaks is a testament to this
intervention, though the frequency of PM\textsubscript{2.5} peaks is
still beyond acceptable.

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM\textsubscript{10} and PM\textsubscript{2.5}
in fig-24-hour-average. As well as dampening local variation and peaks,
the WHO air quality guidelines are also based on 24-hour averages which
are the standard against which the health risks can be judged.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Habitat International}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi},
  pdfauthor={Elizabeth Tilley; Hope Chilunga; Jonathan Kwangulero; Lars Schöbitz; Saloni Vijay; Heiko Heilgendorff; Marc Kalina},
  pdfkeywords={waste management, healthcare waste, trash burning, air
quality, Malawi, urbanisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangulero@gmail.com} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[1]{Saloni Vijay%
%
}
 \ead{svijay@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{heiko.heilgendorff@gmail.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, enhanced, sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus\footnote{Sharps are handled and disposed of separately.}
\citep{tilley2020wea}. However, the vast majority of the deposited waste
is not incinerated, as a new incinerator provided by the Ministry of
Health in late 2019, is only able to handle a small percentage of the
hospital's waste. Furthermore, the incinerator is frequently inoperable,
under repair, or without power\footnote{There was an older incinerator
  on the grounds which failed in 2017. Between 2017 and 2021, when the
  new incinerator was commissioned, nothing was incinerated at all.}.
Rather, the waste is openly burned, in a constantly smouldering pile to
the side of the incinerator building, sending up the aforementioned
clouds of smoke, which choke the grounds
(Figure~\ref{fig-open-burning}). Innumerable other, smaller, fires,
scattered across the QECH campus (Figure~\ref{fig-map-sensors})
contribute their part, as grounds and maintenance staff habitually burn
leaves and other garden refuse. Furthermore, caregivers\footnote{Patients
  are required to bring a caregiver for in-patient hospital stays,
  usually a family member, who assists with feeding, bathing, and all
  other non-medical services.} who reside separately on hospital
grounds, cook for themselves throughout the day and night using biomass
(charcoal or wood), and burn their own domestic waste. As a result, air
quality at the hospital is a constant source of discussion, with
patients, staff, and caregivers struggling to cope, when just breathing
is, as the quote\footnote{ID5, 2019-11-18} used in the title of this
article, `unbearable.'

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trash
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM\textsubscript{2.5}
anthropogenic emissions, it is of particular concern in Sub-Saharan
African (SSA) cities, which are home to 19 of the world's 50 biggest
dumpsites (United Nations Development Programme, 2021). Particulate
matter, or PM, is not specific to trash burning, but rather the product
of any type of incomplete combustion and is one of the most commonly
measured indicators for air quality. PM is measured in terms of its
particle size and generally classified into PM with particle sizes less
than 10 µm (PM\textsubscript{10}) and those with sizes less than 2.5 µm
(PM\textsubscript{2.5}), while one way of measuring the severity of air
pollution is by expressing the mass of PM per volume of air (µg
PMx/m\textsuperscript{3} air). PM has been linked to multiple negative
health outcomes. Amongst particulate air pollution, PM\textsubscript{10}
and PM\textsubscript{2.5} are of particular concern as they can
penetrate deep into the lungs, and their exposure has been associated
with asthma, heart disease, heart failure, stroke, and cancer
\citep{mohapatra2014effect, xing2016impact}. As a result, the health
impacts of ambient air particle pollution may be significant, especially
for the most vulnerable. For instance, the \citet{ihme2019ambient}
estimates PM related air pollution resulted in 4.14 million deaths
worldwide (in 2019), of which, 394,000 occurred in Africa. Yet, despite
its importance, the air quality and health picture in Africa is
worryingly incomplete. For instance, the WHO only collects data from 10
African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

\begin{itemize}
\item
  \textbf{Lions Sight First Eye Hospital}, often shortened to
  \textbf{Lions Sight}, is the largest eye hospital in Malawi and serves
  as the main teaching eye hospital for the Kamuzu University of Health
  Sciences (formerly College of Medicine). It is staffed by 5
  consultants and a team of clinical officers and nurses. It provides a
  mix of public (no cost) and private (at cost) services.
\item
  \textbf{The Blantyre Malaria Project (BMP)}, established by Professors
  Terrie Taylor and Malcolm Molyneux, has carried out clinical research
  and patient care in the area of paediatric malaria, specifically
  cerebral malaria, for more than 30 years\footnote{BMP is credited with
    the development of the Blantyre Coma Score, which is used widely.}.
  Its research infrastructure includes an administrative team, an
  inpatient research ward, an MRI centre, and an outpatient research
  clinic in Ndirande township and in other districts outside of
  Blantyre.
\item
  \textbf{Ward 6B} is a male ward for trauma and orthopaedic patients.
  It was established as part of the original hospital design. It is a
  60-bed ward with a dedicated nursing station and treatment rooms.
\item
  \textbf{The Blantyre Lighthouse Trust Clinic}, also known as
  \textbf{Umodzi Family Centre} or simply, the \textbf{Lighthouse
  Clinic}, is one of four operating across Malawi. The clinics work in
  collaboration with the Ministry of Health (MOH) to provide integrated
  HIV testing, treatment and care for people living with HIV.
\item
  The \textbf{Moyo Nutritional Rehabilitation and Research Unit}, known
  on campus as \textbf{uMoyo}, is a 57-bed nutritional rehabilitation
  unit (NRU) for treating infants and children with severe malnutrition
  and acute illnesses. It is one of 104 operational NRUs in Malawi. It
  also has an outpatient therapeutic feeding program (OTP) for children
  with malnutrition who can be treated outside of the hospital.
\item
  \textbf{Mercy James Institute for Pediatric Surgery and Intensive Care
  (MJC)}, shortened on campus to \textbf{Mercy James}, was opened in
  2017 by the NGO Raising Malawi, which was founded by Madonna. With 3
  operating rooms and 50 beds, it is the first and only first dedicated
  paediatric surgery hospital in Malawi. Prior to its opening, QECH had
  fewer than 10 intensive care beds.
\item
  The \textbf{Administration Building} is the operational hub of the
  campus, containing mostly offices and meeting rooms.
\item
  When a patient arrives at QECH, she or he will usually arrive with a
  guardian: someone to cook for them, buy medicine, do their laundry,
  and help them bathe. The \textbf{Guardian Shelter} is a
  gender-separated concrete floor shelter for sleeping along with a
  cooking pavilion and toilet/shower block that are maintained by a
  local NGO, Chira fund. Though simple, the facility is secure and
  offers one of the only free/accessible toilet facilities for visitors
  on the QECH campus.
\end{itemize}

@@

Despite its national prominence, and international reputation for
academic medical research, Queen's is plagued by constant shortages of
human and material resources. Yet, these shortages are not shared
equally across the space, with a sharp division in cleanliness,
resources, and quality of care between public wards, supported by the
Malawi Ministry of Health, and the private wads with international
funding, such as the aforementioned Mercy James Institute. For staff,
these shortages have manifested in poor salaries, missed pay cheques,
failing infrastructure and limited supplies \citep{tilley2020wea}.
Regarding waste management, these shortages have manifested in a chronic
shortage of black bags, waste containers and cleaning staff. As a
result, due to a lack of bags and storage containers, there is minimal
segregation of waste, and highly infectious items (bandages, gloves,
syringes, and other sharps) are disposed of along with non-hazardous
waste (food, newspapers, packaging). As a result, rather than
facilitating different disposal pathways for hazardous and non-hazardous
waste, all waste is mixed, necessitating more burning than would be
necessary if non-hazardous waste was diverted, and contributing to
workplace risk, with reports of accidental exposure to potentially
infected waste, including accidental pokes from improperly disposed of
sharps, common amongst hospital staff \citep{kalina2020this}.
Regardless, separation ends at the point of collection. The hospital is
not serviced by regular municipal waste collection, and as a result,
nearly all of the waste produced by the hospital, hazardous and
non-hazardous, has to be disposed of on-site, either within the limited
capacity of the hospital incinerator, or more commonly, through open
burning. As a result, the constantly smouldering pile of waste puts
forth a continuous cloud of grey smoke, which mingling with the dozens
of other fires on the grounds, from cooking fires and burning garden
waste, blankets Queen's in a permanent cloud of foul smelling haze.

@@

The PM monitoring device centres around a Raspberry Pi 3 Model B
single-board computer complete with an operating system and storage
space. The added advantage of storage space is that the data are not
lost if the internet connection breaks down. Because the Pi can only
read digital signals, we needed to include an analogue-to-digital (ADC)
converter. The amount of dust is measured by a Nova PM SDS011 High
Precision Laser sensor. This sensor measures particles at 2.5 and 10 µm
in diameter in µg/m\textsuperscript{3}. A list of the specific hardware
and software components can be found in Appendix B.

@@

Raw data and analysis-ready processed data is available as an R package
at \citet{R-qechairquality}. The data underlying the tables and figures
of this manuscript are contained in a repository alongside a
reproducible document that contains the analysis code and the written
narrative of the manuscript \citep{schobitz2023manuscript}.

@@

\hypertarget{tbl-peaks}{}
\begin{longtable}[]{@{}lrrr@{}}
\caption{\label{tbl-peaks}Peaks for both PM\textsubscript{2.5} and
PM\textsubscript{10} are the number of data points above the WHO interim
target 1 (annual) which is the least stringent; the number of
observations recorded (count) is provided for reference.}\tabularnewline
\toprule\noalign{}
location & n & pm10 & pm2.5 \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
location & n & pm10 & pm2.5 \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Administration & 10900 & 171 & 0 \\
Guardian Shelter & 11032 & 4529 & 0 \\
Lighthouse Clinic & 11472 & 4365 & 0 \\
Lions Sight & 11749 & 91 & 0 \\
Malaria & 12031 & 617 & 0 \\
Mercy James & 12193 & 736 & 0 \\
Ward 6B & 11861 & 1113 & 0 \\
uMoyo & 10637 & 703 & 0 \\
\end{longtable}

Over the course of the two-month study, the hazardous limits for both
parameters were exceeded at all locations except for the Lions Sight
hospital. The limits were exceeded fewer than 50 times at five of the
locations, and only the monitors at the Lighthouse Clinic and the
Guardian Shelter recorded more than 50 instances above the hazardous
limits. At the Lighthouse Clinic, the PM\textsubscript{10} limit was
exceeded almost twice as often as the PM\textsubscript{2.5} limit, while
at the Guardian Shelter, the reverse is true. Because the Lighthouse
Clinic is within 50 m of the incinerator, the number and predominance of
PM\textsubscript{10} peaks is characteristic of incomplete combustion
and dust that is typical for the area and the open burning that occurs
there. Although there is trash burning at the Guardian Shelter, the main
source of smoke is from the constant cooking. The residents at the
Guardian Shelter are mostly cooking coal and wood that is burned within
small ``improved'' cooking stoves that are contained, in a row, within a
dedicated cooking block. The fewer number of PM10 peaks is a testament
to this intervention, though the frequency of PM2.5 peaks is still
beyond acceptable.

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM\textsubscript{10} and PM\textsubscript{2.5}
in Figure~\ref{fig-24-hour-average}. As well as dampening local
variation and peaks, the WHO air quality guidelines are also based on
24-hour averages which are the standard against which the health risks
can be judged.
